---
title: "no-pie-charts"
author: "hannarud"
date: "15 марта 2016 г."
output: html_document
---

Following the <http://www.r-bloggers.com/how-to-replace-a-pie-chart/> article content.

# Setup

I start by transcribing the data directly from the plot into R. `readr::read_csv` is useful for constructing a table on the fly:

```{r}
library(readr)

d <- read_csv("Task,< 1 a week,1-4 a week,1-3 a day,>4 a day
Basic exploratory data analysis,11,32,46,12
Data cleaning,19,42,31,7
Machine learning/statistics,34,29,27,10
Creating visualizations,23,41,29,7
Presenting analysis,27,47,20,6
Extract/transform/load,43,32,20,5")

# Reorganize
library(tidyr)
d <- gather(d, Hours, Percentage, -Task)
```

This constructs our data in the form:

```{r}
head(d, 10)
```

# Bar plot

The most common way a pie chart can be improved is by turning it into a bar chart, with categories on the x axis and percentages on the y-axis.

This doesn’t apply to all plots, but it does to this one.

```{r}
library(ggplot2)
positions <- c("< 1 a week", "1-4 a week", "1-3 a day", ">4 a day")
theme_set(theme_bw())
ggplot(d, aes(Hours, Percentage)) +
  geom_bar(stat = "identity") +
  scale_x_discrete(limits = positions) +
  facet_wrap(~Task)
```

# Alternative plots

This was one of a few alternatives I considered when I first imagined creating the plot. When you’ve made a lot of plots, you’ll learn to guess in advance which you will be worth trying, but often it’s worth visualizing a few just to check.

We have three attributes in our data: `Hours`, `Task`, and `Percentage`. We chose to use `x`, `y`, and `facet` to communicate those respectively, but we could have chosen other arrangements. For example, we could have had `Task` represented by color, and represented it with a line plot:

```{r}
ggplot(d, aes(Hours, Percentage, color = Task, group = Task)) + geom_line() + 
  scale_x_discrete(limits = positions)
```

This has some advantages over the above bar chart. For starters, it makes it trivially easy to compare two tasks. (For example, we learn that “Creating visualizations” and “Data cleaning” take about the same distribution of time). I also like how obvious it makes it that “Basic exploratory data analysis” takes up more time than the others. But the graph makes it harder to focus just one one task, you have to look back and forth from the legend, and there’s almost no way we could annotate it with text like the original plot was.

Here’s another combination we could try:

```{r}
ggplot(d, aes(Hours, Task, fill = Percentage)) +
  geom_tile(show.legend = FALSE) +
  geom_text(aes(label = paste0(Percentage, "%")), color = "white") +
  scale_x_discrete(limits = positions)
```

**Note added by hanna**

We can also do it in a reverse: the more percentage, the heavier color:

```{r}
ggplot(d, aes(Hours, Task, fill = 100-Percentage)) +
  geom_tile(show.legend = FALSE) +
  geom_text(aes(label = paste0(Percentage, "%")), color = "white") +
  scale_x_discrete(limits = positions)
```

# Improving our graph

How can our bar plot be improved?

The first problem that jumps out is that the x-axis overlaps so the labels are nearly unreadable. This can be fixed with this solution.

```{r}
ggplot(d, aes(Hours, Percentage)) +
  geom_bar(stat = "identity") +
  facet_wrap(~Task) +
  theme(axis.text.x = element_text(angle = 90,  hjust = 1)) +
  scale_x_discrete(limits = positions)
```

Next, note that the original pie chart showed the percentages as text right on the graph. This was necessary in the pie chart simply because it’s so difficult to guess a percentage out of a pie chart- we could afford to lose it here, when the y-axis communicates the same information. But it can still be useful when you want to pick out a specific number to report (“Visualization is important: 7% of data scientists spend >4 hours a day on it!”) So I add a `geom_text` layer.

```{r}
ggplot(d, aes(Hours, Percentage)) +
  geom_bar(stat = "identity") +
  facet_wrap(~Task) +
  geom_text(aes(label = paste0(Percentage, "%"), y = Percentage),
            vjust = 1.4, size = 5, color = "white") +
  scale_x_discrete(limits = positions)
```

The ordering of task facets is arbitrary (alphabetical in this plot). I like to give them an order that makes them easier to browse- something along the lines of. A simple proxy for this is to order by “% who spend < 1 hour a week.”

```{r}
library(dplyr)

d %>%
  mutate(Task = reorder(Task, Percentage, function(e) e[1])) %>%
  ggplot(aes(Hours, Percentage)) +
  geom_bar(stat = "identity") +
  facet_wrap(~Task) +
  geom_text(aes(label = paste0(Percentage, "%"), y = Percentage),
            vjust = 1.4, size = 5, color = "white") +
  theme(axis.text.x = element_text(angle = 90,  hjust = 1)) +
  xlab("Hours spent per week") + scale_x_discrete(limits = positions)
```

# Graph design

From here, the last step would be to adjust the colors, fonts, and other “design” choices.

I don’t have terribly strong opinions about these choices (I’m pretty happy with ggplot2’s `theme_bw()`). But some prefer Edward Tufte’s approach of maximizing the “Data/Ink Ratio”- that is, dropping borders, grids, and axis lines. This can be achieved with `theme_tufte`:

```{r}
library(ggthemes)

d %>%
  mutate(Task = reorder(Task, Percentage, function(e) e[1])) %>%
  ggplot(aes(Hours, Percentage)) +
  geom_bar(stat = "identity") +
  facet_wrap(~Task) +
  geom_text(aes(label = paste0(Percentage, "%"), y = Percentage),
            vjust = 1.4, size = 5, color = "white") +
  theme_tufte() +
  theme(axis.text.x = element_text(angle = 90,  hjust = 1)) +
  scale_x_discrete(limits = positions)
```

Some people take this philosophy even further, and drop the y-axis altogether (since we do already have those percentages annotated on the bars).

```{r}
d %>%
  mutate(Task = reorder(Task, Percentage, function(e) e[1])) %>%
  ggplot(aes(Hours, Percentage)) +
  geom_bar(stat = "identity") +
  facet_wrap(~Task) +
  geom_text(aes(label = paste0(Percentage, "%"), y = Percentage),
            vjust = 1.4, size = 5, color = "white") +
  theme_tufte() +
  theme(axis.text.x = element_text(angle = 90,  hjust = 1),
        axis.ticks = element_blank(),
        axis.text.y = element_blank()) +
  ylab("") +
  scale_x_discrete(limits = positions)
```

# From comments

## What's wrong with a stacked bar plot, the area-preserving cousin to the pie chart?

```{r}
d %>%
mutate(Hours = factor(Hours, levels = unique(d$Hours))) %>%
ggplot(aes(x=Task,y=Percentage)) + geom_bar(stat="identity", position="stack", aes(fill=Hours)) +
  coord_flip() +
  scale_fill_brewer(palette="YlGnBu") +
  theme_minimal() +
  theme(legend.position="bottom")
```

## I think I would take a different approach, as you allude to in your footnote, and just display a single bar chart showing the "average" number of hours per day that the respondents spend on various tasks.

```{r}
# convert the categories reported to some middle value of hours per day
hrsperday <- c(0.1, 0.4, 2.5, 6)

# use these as weights to calculate the weighted mean for each task
totals <- apply(t(d[, 3]/100)*hrsperday, 2, sum)
tot <- data.frame(taskname = d[, 1], tasktime = totals)

# sort the tasks by time spent
totals <- tot[order(tot$tasktime),]

# plot the results
par(mar=c(4, 13, 0, 1))
ggplot(totals, aes(Task, tasktime)) + geom_bar(stat = "identity")
```
